---
title: " HW Question Week 1 "
subtitle: "STA2101F 2021"
author: "Amy Farrow"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
fontsize: 12pt
---

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  out.width = "100%",
  fig.width = 10,
  fig.height = 6.7, 
  fig.retina = 3,
  cache = FALSE)
```

```{r library}
library(faraway)
data(wafer)
library(tidyverse)
library(modelsummary)
library(kableExtra)
```

**(a) Fit the linear model `resist ~ x1 + x2 + x3 + x4`. Extract the $X$ matrix using the `model.matrix` function. How have the levels of the factors been coded?**

Model 1:

```{r}
model1 <- lm(resist ~ x1 + x2 + x3 + x4,
             wafer)

sumary(model1)
```

$X$ Matrix:

```{r}
(X <- model.matrix(model1)) %>%
  kable()
```

\- is coded as 0 and + is coded as 1.

**(b) Compute the correlation between the columns of the $X$ matrix. Why are there some missing values?**

Correlation Matrix:

```{r}
cor(X) %>%
  kable()
```

The standard deviation of the Intercept column in $X$ is zero, so the correlation coefficient between it and all other columns is not calculable.

**(c) What difference in resistance is expected when moving from the low to the high level of `x1`?**

From part (a), we can expect an increase of 25.762 when moving from the low to the high level of `x1`.

**(d) Refit the model without `x4` and examine the regression coefficients and standard errors. What stayed the same and what changed? How is this related to the correlation matrix of $X$?**

Model 2:

```{r}
model2 <- lm(resist ~ x1 + x2 + x3,
             wafer)

sumary(model2)
```

Comparing to part (a), the estimated coefficients for `x1`, `x2`, and `x3` do not change. This is expected because the correlation matrix of $X$ shows zero correlation between each pair of `xi`.

The standard errors, t-values, and p-values for `xi` change slightly (increasing, magnitude decreasing, and decreasing, respectively). The intercept estimate decreases slightly, and the corresponding standard error decreases, t-value increases in magnitude, and p-value decreases.
